---
title: "Alignment Play!"
output: 
  html_notebook:
    toc: true
---


# Introduction

I have put a teeny little "genome", which is basically 60 Kb of non-repetitive and non-missing
DNA from Chinook salmon
and put it into the file `genome/tiny.fna`. It is a fasta file.  I have some code below
that prepares that "genome" (indexes it) so that you can align sequences against it
with `bwa mem`.  I then have written some R functions to automate the process of
taking two sequences (interpreted as paired-end reads) and mapping them to our tiny genome
and then printing the relevant lines from the SAM file for your perusal.

We build up these tools so that you can quickly just make up any two sequences you
want to map them to the genome.  Then we are going to ask you to create all sorts of
different sequences to obtain particular results (mapping posiitons, flags, CIGAR strings, etc.).
So, get ready!

# Indexing the genome, and slurping it up

For fun, look at the first 10 lines of the fasta file.
```{bash}
head genome/tiny.fna
```
This "genome" only has one scaffold in it.  This is atypical.  Most genomes are in many
different pieces.
Before we can map sequences against this silly, fake, tiny single-scaffold genome with `bwa`
we must _index_ it with `bwa`.  That is super easy:
```{sh}
bwa index genome/tiny.fna
```

## Now, store the sequence in a string

In order to grab little pieces from known parts of that genome,
will will put it into an R object as a string:
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
gen_seq <- read_lines("genome/tiny.fna")[-1] %>%
  paste(collapse = "")
```

Now, if you want to get a 100 bp segment from (base-1 indexed) positions 2001 to 2100,
and another from 2401 to 2500,
you can do it like this:
```{r}
R1 <- str_sub(gen_seq, 2001, 2100)
R2 <- str_sub(gen_seq, 2401, 2500)
```

# A function to get SAM output from mapping pairs of sequences

Here we just write a function that takes as input two sequences, say R1 and R2, and then
writes them into two fastq files and maps them as paired reads using `bwa mem`, and then
spits out the resulting SAM file to stdout.
```{r}
#' map two reads with bwa mem and print the alignment lines
#' @param R1 sequence of read 1 of the pair
#' @param R2 sequence of read 2 of the pair.
map_two_reads <- function(R1, R2) {
  
  # make a "workspace" directory to do this in and ensure that it is clean
  dir.create("workspace", showWarnings = FALSE)
  dump <- file.remove(dir("workspace/", full.names = TRUE))
  
  # create two fastq files with the sequences in there.
  # given them super high quality scores
  # here is the first seq
  cat("@FAKESEQ/1", 
      R1, 
      "+", 
      paste(rep("J", nchar(R1)), collapse = ""), 
      sep = "\n", 
      file = "workspace/seq_R1.fq")
  
  # and here is the second
  cat("@FAKESEQ/2", 
      R2, 
      "+", 
      paste(rep("J", nchar(R2)), collapse = ""), 
      sep = "\n", 
      file = "workspace/seq_R2.fq")
  
  # now map them with bwa mem and print the resulting SAM without the header
  system("cd workspace; bwa mem ../genome/tiny.fna seq_R1.fq seq_R2.fq 2> bwa.stderr")
}
```

Let's demonstrate its use.  Note that the output here is going to get wrapped in the output text box,
but that is OK. It is easy to see the different columns and the different lines:
```{r}
map_two_reads(R1, R2)
```

OK! That is quick and easy, and let's us play around with modifying sequences to let use see the
results on the mapping.

Before going on, let's just point out that if you want to know what the SAM flags mean, one
of the simplest ways of figuring that out is to use the Broad Institute's interactive
web site at
[https://broadinstitute.github.io/picard/explain-flags.html](https://broadinstitute.github.io/picard/explain-flags.html).

# One more convenient function

Pass this function a DNA string and it will return its reverse complement:
```{r}
revcomp <- function(R) {
  cmp <- c(A = "T", C = "G", G = "C", T = "A")  # vector of bases named as their complements
  paste(rev(cmp[strsplit(R, "")[[1]]]), collapse = "")
}
```

Now, we can see what the alignment looks like if the template had been the
complementary strand:
```{r}
map_two_reads(revcomp(R1), revcomp(R2))
```

Note that the only thing that has changed are the SAM flags.